Temperature Effects on Surface Sulfur Dioxide

### 2011-07-21 Background.tex
Finished writing out the hand-notes for coordinations + rdfs, and the background for coordinations and cycles

A hydrated \suldiox~in an aqueous environment forms hydrogen bonds through the oxygens to nearby water hydrogens, or interacts via the sulfur atoms with water oxygens. To further our analysis of the way in which \suldiox~coordinates its bonding to surface waters, we adopt a naming scheme to denote the way in which the \suldiox~is hydrated by the surrounding waters. This naming scheme mimics a notational system developed during a previous study on water coordination by Buch et al,\cite{Buch 2005} and was subsequently used in other computational work.\cite{Walker2006b} In this naming system, a letter is used to designate the atom on a water molecule through which a hydrogen bond is formed to neighboring waters. Thus a bonding coordination of ``OOH'' designates two proton-accepting bondind interactions through the water oxygen, and a single proton-donor bonding interaction through a hydrogen. More recently, Baer et al devised a nomenclature that explicitly enumerates the bonding to \suldiox~via the sulfur or oxygen atoms.\cite{Baer2010}

In this work we port Buch et al's nomenclature to \suldiox~in order to show the amount of hydrogen bonding through the acceptor \suldiox-oxygens, and the weaker bonding interactions form from the \suldiox-sulfur to water oxygens. Thus, an ``SOO'' coordinated \suldiox~molecule forms a single iteraction through the sulfur atoom to a neighboring water oxygen, and two hydrogen bonds through either a single \suldiox-oxygen, or distributed with one hydrogen bond on each of the \suldiox~oxygens. Analysis of the distribution of \suldiox~coordinations will give insight to how it binds to the water surface.

@@

Hydrated \suldiox~clusters have been studied extensively with many recent experiments and computations congealing to a clearer picture of \suldiox~bulk and surface behaviors.\cite{Baer2010, Tarbuck2005, Tarbuck2006, Ota2011, Bishenden1998, Hirabayashi2006, Steudel2009, Yang2002, Hayash1985, Moin2011, Eckl2008} At a water surface, it is now known that \suldiox~forms a complex with water during adsorption, and then subsequently absorbs into the interfacial region by reaction to form ionic sulfur species.\cite{Tarbuck2005, Tarbuck2006, Ota2011} The first attempt to elucidate the structure of the surface hydrated \suldiox~by Baer et al. showed that the bonding coordination distributions of \suldiox~at the surface are altered relative to the bulk region, and that two coordinations dominate the distribution of bonding types: the ``SO'' and the ``SOO''. In the same work they then focused on the moste dominant coordination to determine the most likely cluster geometry using two and three-water clusters bound to the \suldiox.

@@

Here we will only briefly introduce graph theoretical concepts as they have been described well by others, and with varied application to cyclic structures.\cite{Tutte1984, Balakrishnan2000, Harary1973, Huber2007, Garcia2004, Dury2001} A graph consists of a set of vertexes, and edges that connect the vertexes. A molecule can be represented with atoms as vertexes, and edges for each intramolecular covalent bond connecting the atoms. The set of edges is then further expanded to include intermolecular interactions such as hydrogen bonds and other bonding bonding interactions. Edges may be assigned weights (i.e. bondlengths), types, and can be directional pointing towards a specific target from a source. A molecular system including all atoms, bonds, and interactions is thus fully described by a graph.

To detect cyclic structures in a graph a depth-first or breadth-first search (DFS and BFS, respectively) may be used.\cite{Knuth1997, Cormen 2001} A BFS is a recursive algorithm of queueing vertexes and all adjacent neighboring vertexes while performing a specified procedure on each visited vertex. This is easily performed on adjaceny list or connectivity matrix data structures, iterating through atms of interest in the graph as starting points of the search. In BFS terminology, all nodes are colored during graph traversal to distinguish unvisited nodes (white), queued nodes (grey), and visited nodes (black). Using a BFS on a graph, cyclic structures are detected any time a ``grey target'' is encountered when queueing adjacent neighbors of a vertex. A benefit of BFS on a graph is the ability to determine the smallest cyclic structure in which a given vertex is a member. In the case of \suldiox~hydrate structures, beginning the BFS with the \suldiox~sulfur as the root vertex will discover cyclic bonding structures in order of size. Here we are only concerned with the smallest cyclic structure involving those waters in the first and second hydration shells aroun the \suldiox. Furthermore, it is possible to reconstruct a cycle's structure to find the size of the cycle (number of atoms involved), the number of unique waters in the cycle, and the type of cyclic structure encountered.

@@

Baer et al. presented a detailed geometric and spectroscopic breakdown of types \Rmnum{3} cycles with two and three waters from their DFT calculations.\cite{Baer2010} Given the information of the number of waters, atoms, and bonds involved in the bonding cycles, we find that of the three-water type \Rmnum{3} cycles, there exist two structural varietes, shown in Figure \ref{fig:type-3-varieties}, that differ in the arrangement of the waters involved. Type \Rmnum{3}A is arranged with each water contributing an OH bond to the structure of the cycle. Type \Rmnum{3}B involves a single water contributing an OH, whereas the other two waters contribute only the oxygen atom or the entire water molecule to the structure, respectively. This nuance of the three-water type \Rmnum{3} structures, and the overall distribution are presented in more detail later. We also show the distribution of cyclic structures encountered during MD simulations to further understand the behaviors of \suldiox-hydrates at the water surface.

>>>>

A hydrated \suldiox~in an aqueous environment forms hydrogen bonds through the oxygens to nearby water hydrogens, or interacts via the sulfur atoms with water oxygens. To further our analysis of the way in which \suldiox~coordinates its bonding to surface waters, we adopt a naming scheme to denote the way in which the \suldiox~is hydrated by the surrounding waters. This naming scheme mimics a notational system developed during a previous study on water coordination by Buch et al,\cite{Buch 2005} and was subsequently used in other computational work.\cite{Walker2006b} In this naming system, a letter is used to designate the atom on a water molecule through which a hydrogen bond is formed to neighboring waters. Thus a bonding coordination of ``OOH'' designates two proton-accepting bonding interactions through the water oxygen, and a single proton-donor bonding interaction through a hydrogen. More recently, Baer et al devised a nomenclature that explicitly enumerates the bonding to \suldiox~via the sulfur or oxygen atoms.\cite{Baer2010}

In this work we port Buch et al's nomenclature to \suldiox~in order to show the amount of hydrogen bonding through the acceptor \suldiox-oxygens, and the weaker bonding interactions form from the \suldiox-sulfur to water oxygens. Thus, an ``SOO'' coordinated \suldiox~molecule forms a single interaction through the sulfur atom to a neighboring water oxygen, and two hydrogen bonds through either a single \suldiox-oxygen, or distributed with one hydrogen bond on each of the \suldiox~oxygens. Analysis of the distribution of \suldiox~coordinations will give insight to how it binds to the water surface.

@@

Hydrated \suldiox~clusters have been studied extensively with many recent experiments and computations congealing to a clearer picture of \suldiox~bulk and surface behaviors.\cite{Baer2010, Tarbuck2005, Tarbuck2006, Ota2011, Bishenden1998, Hirabayashi2006, Steudel2009, Yang2002, Hayashi1985, Moin2011, Eckl2008} At a water surface, it is now known that \suldiox~forms a complex with water during adsorption, and then subsequently absorbs into the interfacial region by reaction to form ionic sulfur species.\cite{Tarbuck2005, Tarbuck2006, Ota2011} The first attempt to elucidate the structure of the surface hydrated \suldiox~by Baer et al. showed that the bonding coordination distributions of \suldiox~at the surface are altered relative to the bulk region, and that two coordinations dominate the distribution of bonding types: the ``SO'' and the ``SOO''. In the same work they then focused on the most dominant coordination to determine the most likely cluster geometry using two and three-water clusters bound to the \suldiox.

@@

Here we will only briefly introduce graph theoretical concepts as they have been described well by others, and with varied application to cyclic structures.\cite{Tutte1984, Balakrishnan2000, Harary1973, Huber2007, Garcia2004, Dury2001} A graph consists of a set of vertexes, and edges that connect the vertexes. A molecule can be represented with atoms as vertexes, and edges for each intramolecular covalent bond connecting the atoms. The set of edges is then further expanded to include intermolecular interactions such as hydrogen bonds and other bonding bonding interactions. Edges may be assigned weights (i.e. bond lengths), types, and can be directional pointing towards a specific target from a source. A molecular system including all atoms, bonds, and interactions is thus fully described by a graph.

To detect cyclic structures in a graph a depth-first or breadth-first search (DFS and BFS, respectively) may be used.\cite{Knuth1997, Cormen 2001} A BFS is a recursive algorithm of queuing vertexes and all adjacent neighboring vertexes while performing a specified procedure on each visited vertex. This is easily performed on adjacency list or connectivity matrix data structures, iterating through atoms of interest in the graph as starting points of the search. In BFS terminology, all nodes are colored during graph traversal to distinguish unvisited nodes (white), queued nodes (gray), and visited nodes (black). Using a BFS on a graph, cyclic structures are detected any time a ``gray target'' is encountered when queuing adjacent neighbors of a vertex. A benefit of BFS on a graph is the ability to determine the smallest cyclic structure in which a given vertex is a member. In the case of \suldiox~hydrate structures, beginning the BFS with the \suldiox~sulfur as the root vertex will discover cyclic bonding structures in order of size. Here we are only concerned with the smallest cyclic structure involving those waters in the first and second hydration shells around the \suldiox. Furthermore, it is possible to reconstruct a cycle's structure to find the size of the cycle (number of atoms involved), the number of unique waters in the cycle, and the type of cyclic structure encountered.

@@

Baer et al. presented a detailed geometric and spectroscopic breakdown of types \Rmnum{3} cycles with two and three waters from their DFT calculations.\cite{Baer2010} Given the information of the number of waters, atoms, and bonds involved in the bonding cycles, we find that of the three-water type \Rmnum{3} cycles, there exist two structural varieties, shown in Figure \ref{fig:type-3-varieties}, that differ in the arrangement of the waters involved. Type \Rmnum{3}A is arranged with each water contributing an OH bond to the structure of the cycle. Type \Rmnum{3}B involves a single water contributing an OH, whereas the other two waters contribute only the oxygen atom or the entire water molecule to the structure, respectively. This nuance of the three-water type \Rmnum{3} structures, and the overall distribution are presented in more detail later. We also show the distribution of cyclic structures encountered during MD simulations to further understand the behaviors of \suldiox-hydrates at the water surface.

### 2011-07-21 Coordination-Results.tex
Finished writing out the hand-notes for coordinations + rdfs, and the background for coordinations and cycles

We thus have a likely route for adsorption and binding of an unbound \suldiox~to a water surface. The unbound \suldiox~initially binds mostly to a water through the sulfur atom, and then will either unbind, or add an oxygen interaction to oscillate between unbound, ``S'', and ``SO'' coordinations. It may then undergo brief sojourns to other coordinations because of the motions and rearrangements of the surface water molecules. These brief transitory coordinations are overwhelmed by the dominant ``S'' and ``SO'' states. The most probable deviation from those two coordinations is to the ``SOO'', where a second oxygen interaction is formed. Otherwsie the \suldiox~unbinds from the surface to repeat the adsorption process.

Baer et al performed this coordination analysis for their single-temperature study, but discriminating between \suldiox~binding through the two different oxygens. They concluded that there is assymetric hydrogen bonding through the \suldiox~oxygens, with one binding more often than the other. This is supported by the findings here where all the double oxygen coordinations (i.e. ``OO'', ``SOO'', etc.) represent a much lower percentage of the coordinations than the single oxygen counterparts (i.e. ``O'', ``SO'', etc.).

@@

% Figure showing all the different types of so2 coordinations
\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/coordination/so2-bonding-coordinations.png}
		\caption{Distribution of bonding coordinations of the surface \suldiox~molecule. Each set of bars represents the amount of time spent in the particular coordination type, including the unbound coordination. The results for the cold (blue) and hot (red) temperatures are shown together. At the high temperature the \suldiox~bonding becomes more complex with more bonds to the sulfur and oxygen atoms. The lower coordination states are more dominated in the cold temperature. The naming scheme is similar to that of a previous scheme for water coordinations,\cite{Buch2005} applied to the \suldiox~molecule.}
		\label{fig:so2-bonding-coordinations}
	\end{center}
\end{figure}

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/coordination/coordination-breakdown.png}
		\caption{Bonding coordinations of the \suldiox~broken down into specific interactions. The total number of bonds to the molecule (top) are shown regardless of coordination type. The total number of bonds through either the sulfur atom (center) or through both of the oxygens (bottom). The \suldiox~primarily forms one or two bonds to the surface waters. The sulfur atom mostly bonds to a single water oxygen. Between the two \suldiox~oxygen atoms, the molecule mostly does not bond to the water hydrogens, or forms a single hydrogen-bond.}
		\label{fig:coordination-breakdown}
	\end{center}
\end{figure}

@@

These increased

>>>>

We thus have a likely route for adsorption and binding of an unbound \suldiox~to a water surface. The unbound \suldiox~initially binds mostly to a water through the sulfur atom, and then will either unbind, or add an oxygen interaction to oscillate between unbound, ``S'', and ``SO'' coordinations. It may then undergo brief sojourns to other coordinations because of the motions and rearrangements of the surface water molecules. These brief transitory coordinations are overwhelmed by the dominant ``S'' and ``SO'' states. The most probable deviation from those two coordinations is to the ``SOO'', where a second oxygen interaction is formed. Otherwise the \suldiox~unbinds from the surface to repeat the adsorption process.

Baer et al performed this coordination analysis for their single-temperature study, but discriminating between \suldiox~binding through the two different oxygens. They concluded that there is asymmetric hydrogen bonding through the \suldiox~oxygens, with one binding more often than the other. This is supported by the findings here where all the double oxygen coordinations (i.e. ``OO'', ``SOO'', etc.) represent a much lower percentage of the coordinations than the single oxygen counterparts (i.e. ``O'', ``SO'', etc.).

@@

% Figure showing all the different types of so2 coordinations
%\begin{figure}[h!]
	%\begin{center}
		%\includegraphics[scale=1.0]{images/coordination/so2-bonding-coordinations.png}
		%\caption{Distribution of bonding coordinations of the surface \suldiox~molecule. Each set of bars represents the amount of time spent in the particular coordination type, including the unbound coordination. The results for the cold (blue) and hot (red) temperatures are shown together. At the high temperature the \suldiox~bonding becomes more complex with more bonds to the sulfur and oxygen atoms. The lower coordination states are more dominated in the cold temperature. The naming scheme is similar to that of a previous scheme for water coordinations,\cite{Buch2005} applied to the \suldiox~molecule.}
		%\label{fig:so2-bonding-coordinations}
	%\end{center}
%\end{figure}

%\begin{figure}[h!]
	%\begin{center}
		%\includegraphics[scale=1.0]{images/coordination/coordination-breakdown.png}
		%\caption{Bonding coordinations of the \suldiox~broken down into specific interactions. The total number of bonds to the molecule (top) are shown regardless of coordination type. The total number of bonds through either the sulfur atom (center) or through both of the oxygens (bottom). The \suldiox~primarily forms one or two bonds to the surface waters. The sulfur atom mostly bonds to a single water oxygen. Between the two \suldiox~oxygen atoms, the molecule mostly does not bond to the water hydrogens, or forms a single hydrogen-bond.}
		%\label{fig:coordination-breakdown}
	%\end{center}
%\end{figure}

@@

These increased binding through the \suldiox~oxygens at the hot temperatures should have a visible effect on the radial distribution function (RDF) between \suldiox-oxygen and water-hydrogen. Figure \ref{fig:rdf} shows the RDF of the \suldiox~sulfur and oxygen atoms with their water bonding partners, water oxygen and hydrogen, respectively, at both cold and hot temperatures. Both the first and second peaks of the O$_{SO_2}$-H$_{H_2O}$ RDF increase at the higher temperature, indicating a large number of water hydrogens near the \suldiox~oxygens in both hydration shells. This corroborates the increased coordination through the \suldiox~oxygens to the surface waters at the higher temperature.

The populations of the single-oxygen coordinations are similar at both temperatures, somewhat contradicting the results of the RDF. We expect an increase of all \suldiox-oxygen interactions with surface water hydrogens, but the non-sulfur coordinations do not obey this trend. This indicates that the non-sulfur coordinations do not behave as most of the other coordinations, and their small contribution to the overall population in the distribution of Figure \ref{fig:bonding-coordinations} indicates that they are anomalous, or only briefly encountered during the simulations.

The same temperature effect of shifting population to higher coordinations also manifests in the \suldiox~sulfur. The single-``S'' coordinations have similar populations at both temperatures, but the change from cold to hot temperature shifts the unbound \suldiox~coordinations up towards to the doubly-bound sulfur coordinations (i.e. ``SS'', ``SSO'', etc.). The increased temperature causes more binding to the \suldiox~by allowing for multiple bonds through both the \suldiox~sulfur, and through the oxygens. Thus, the hotter temperature allows for increased coordination of the \suldiox~to the water surface.

When the \suldiox~is on the cold water surface, the ``S'' and ``SO'' coordination populations are nearly equally represented. As the temperature increases to the hot system, the ``S'' coordination decreases, while the ``SO'' increases. We see from the RDFs that colder temperatures favor binding through the \suldiox~sulfur because of the larger first peak in the S$_{SO_2}$-O$_{H_2O}$ RDF. From this increase we expect the ``S''-coordinated \suldiox~to be more prevalent at cold temperatures, as is the case for the coordination distribution. However, the O$_{SO_2}$-H$_{H_2O}$ RDF indicates increased hydrogen bonding through the \suldiox~oxygens at higher temperatures manifested by a larger first peak in Figure \ref{fig:rdf} for the hot temperature. Thus the minimally ``SO''-coordinated distributions (those with at least a single bond to an oxygen and the sulfur, i.e. ``SO'', ``SSO'', ``SSOO'', etc.) are more prevalent at the high temperature. Overall, we find evidence from both the bonding coordination distribution, and from the RDFs to suggest a modified bonding behavior at different temperatures. At an increased temperature the surface \suldiox~binds more through its oxygens than the cold counterpart. However, like the \suldiox~in the cold temperature, it likely has a persistent interaction through the sulfur atom to a neighboring water oxygen throughout most of the simulated trajectories.
